---
title: "Database Migration: TB"
author:
- Stephen Jones
date: "April 20, 2019"
output:
  rmdformats::readthedown:
    gallery: no
    highlight: tango
    lightbox: yes
    self_contained: yes
    thumbnails: yes
    code_folding: hide
  html_document:
    df_print: paged
---  

<style type="text/css">
pre {
  max-height: 150px;
  float: left;
  width: 100%;
  overflow-y: auto;
}
pre.r {
max-height: none;
}
h1.title {
  color: DarkBlue;
  font-weight: bold;
}
h1 { /* Header 1 */
  color: DarkBlue;
  font-weight: bold;
}
h2 { /* Header 2 */
  color: DarkBlue;
  font-weight: bold;
}
h3 { /* Header 3 */
  color: DarkBlue;
  font-weight: bold;
}
h4 { /* Header 3 */
  color: DarkBlue;
  font-weight: bold;
}
</style>
#Preparation  

First, all items are removed from the workspace. Packages `RMySQL`, `DBI` are loaded to query the local copy of the tuberculosis database created earlier in the semester. The working directory is set to `C:/MSDS/Neo4jDesktop` and the table written to .csv; this path must exist to run the following code chunk.

```{r message=FALSE,warning=FALSE,eval=FALSE}

rm(list=ls())
#grab data from localhost
library(RMySQL)

# Load the DBI library
library(DBI)
# configure connection to MySQL
getSqlConnection <- function() {
  con <-dbConnect(RMySQL::MySQL(),
                  username = 'root',#other ids set up are 'achan' and 'mhayes'
                  password = '1NeWpW1NeWpW',#we all can use the same password
                  host = 'localhost',#this is the IP address of the cloud instance
                  dbname = 'tb')
  return(con)
}

#n=-1 is necessary to grab all records instead of default 500.
connection <- getSqlConnection()
reqst <- dbSendQuery(connection,"select * from tb")
tb_data <- dbFetch(reqst, n=-1)

#working directory is set
setwd('C:/MSDS/Neo4jDesktop')
write.csv(tb_data,"tb_data.csv",row.names=TRUE,na="")

```

The file `tb_data.csv` was uploaded to github (https://raw.githubusercontent.com/sigmasigmaiota/graph/master/tb_data.csv) and downloaded again to prepare for migration to a Neo4j database. The package `dplyr` is used; rownumbers are omitted below.  

```{r message=FALSE,warning=FALSE}

library(dplyr)

file<-("https://raw.githubusercontent.com/sigmasigmaiota/graph/master/tb_data.csv")
file2<-file(file,open="r")
tb<-read.csv(file2,header = TRUE,stringsAsFactors = FALSE)
tb[,1]<-NULL
tb<-na.omit(tb)

```

The password and id for the database are recorded in a hidden code chunk.  

```{r message=FALSE,warning=FALSE,echo=FALSE}

pw<-"NewPassword"
id<-"neo4j"

```  

Using the package `RNeo4j`, a connection is initiated with the database.  

```{r message=FALSE,warning=FALSE}

library(RNeo4j)

#initiate connection to the database
graphtb <- startGraph("http://localhost:7474/db/data", username=id, password=pw)

```

```{r  message=FALSE,warning=FALSE}

#create a query that establishes structure upon which relationships will be detailed
query = "
MERGE (:country {country:{country},year:{year},sex:{sex},child:{child},adult:{adult},elderly:{elderly}})
MERGE (:year {year:{year},sex:{sex},child:{child},adult:{adult},elderly:{elderly},country:{country}})
MERGE (:sex {sex:{sex},child:{child},adult:{adult},elderly:{elderly},country:{country},year:{year}})
MERGE (:child {child:{child},adult:{adult},elderly:{elderly},year:{year},country:{country},year:{year}})
MERGE (:adult {adult:{adult},child:{child},elderly:{elderly},year:{year},country:{country},year:{year}})
MERGE (:elderly {elderly:{elderly},adult:{adult},child:{child},year:{year},country:{country},year:{year}})
"
#create a transaction
tx <- newTransaction(graphtb)

#Upload data to the database row by row.
for (i in 1:nrow(tb)) {
  # Upload in blocks of 1000.
  if(i %% 1000 == 0) {
    # Commit current transaction.
    commit(tx)
    # Call transaction.
    tx = newTransaction(graphtb)
  }

  # implement the query structure detailed above.
  appendCypher(tx,
               query,
               country = tb$country[i],
               year = tb$year[i],
               sex = tb$sex[i],
               child = tb$child[i],
               adult = tb$adult[i],
               elderly = tb$elderly[i])
}
#finish with one laset command.
commit(tx)

```

With the data uploaded to the Neo4j database, relationships must be defined. I define country-to-year relationship (`c_year`), year-to-male adult (`m_adult`), year-to-female adult (`f_adult`), and so on until the relationships in the dataframe are fully described.  

```{r message=FALSE,warning=FALSE}

#country to year
query2<-"MATCH (a:country),(b:year) WHERE (a.year) = (b.year) CREATE (a)-[:c_year]->(b)"
tx2 = newTransaction(graphtb)
appendCypher(tx2, query2)
commit(tx2)

#year to cases, male adults
query3<-"MATCH (a:year),(b:adult) WHERE (a.year) = (b.year) AND (a.sex) = 'male' CREATE (a)-[:m_adult]->(b)"
tx3 = newTransaction(graphtb)
appendCypher(tx3, query3)
commit(tx3)

#year to cases, female adults
query4<-"MATCH (a:year),(b:adult) WHERE (a.year) = (b.year) AND (a.sex) = 'female' CREATE (a)-[:f_adult]->(b)"
tx4 = newTransaction(graphtb)
appendCypher(tx4, query4)
commit(tx4)

#year to cases, male children
query5<-"MATCH (a:year),(b:child) WHERE (a.year) = (b.year) AND (a.sex) = 'male' CREATE (a)-[:m_child]->(b)"
tx5 = newTransaction(graphtb)
appendCypher(tx5, query5)
commit(tx5)

#year to cases, female children
query6<-"MATCH (a:year),(b:child) WHERE (a.year) = (b.year) AND (a.sex) = 'female' CREATE (a)-[:f_child]->(b)"
tx6 = newTransaction(graphtb)
appendCypher(tx6, query6)
commit(tx6)

#year to cases, male elderly
query7<-"MATCH (a:year),(b:elderly) WHERE (a.year) = (b.year) AND (a.sex) = 'male' CREATE (a)-[:m_eld]->(b)"
tx7 = newTransaction(graphtb)
appendCypher(tx7, query7)
commit(tx7)

#year to cases, female elderly
query8<-"MATCH (a:year),(b:elderly) WHERE (a.year) = (b.year) AND (a.sex) = 'female' CREATE (a)-[:f_eld]->(b)"
tx8 = newTransaction(graphtb)
appendCypher(tx8, query8)
commit(tx8)

```

A quick check in the Neo4j browser window confirms the relationships have been set.  

#Query: U.S. and Mexico  

Data for the U.S. and Mexico can provide a glimpse into the prevalence of tb in North America in the past few decades. In the query below, adults, children, and the elderly are called; the number of cases will reflect a sum of all three demographic groups. The package `DT` displays the resulting dataframe `qcountry`.  


```{r message=FALSE,warning=FALSE}

#query for United States and Mexico; all cases are totaled in the last command.
query <- "MATCH (a:country)-->(b:year)
WHERE a.country = 'United States of America' 
OR a.country = 'Mexico' 
AND a.country = b.country AND a.year = b.year
RETURN a.country AS country, SUM(DISTINCT a.adult+a.child+a.elderly) AS cases, b.year AS year"

#data frame stores results
qcountry <- cypher(graphtb, query)

library(DT)
#view results
datatable(qcountry)

```

#Plot: U.S. and Mexico  

`igraph` and `visNetwork` are used to plot the data. Two dataframes are created; in `edges` I define which nodes are to be associated, while in `nodes` I list every node to be included. Each dataframe is derived from `qcountry`.  

```{r message=FALSE,warning=FALSE}

library(igraph)
library(visNetwork)


edges1<-qcountry[c("country","cases")]
colnames(edges1)<-c("from","to")
edges2<-qcountry[c("cases","year")]
colnames(edges2)<-c("from","to")
edges<-data.frame(rbind(edges1,edges2))

nodes<-data.frame(id=unique(c(qcountry$country,qcountry$year,qcountry$cases)),
                  font.size=32)
nodes$label<-nodes$id
nodes$font.size<-ifelse(nodes$id=="United States of America" | nodes$id=="Mexico",48,24)
nodes$level[nodes$id=="United States of America"]<-1
nodes$level[nodes$id>=1995 & nodes$id <= 2013]<-3
nodes$level[nodes$id=="Mexico"]<-5
nodes$level[nodes$id==6934|nodes$id==6623|nodes$id==6274|nodes$id==5884|nodes$id==5651|nodes$id==5434|
              nodes$id==5365|nodes$id==5251|nodes$id==5089|nodes$id==13767|nodes$id==13283|
              nodes$id==12893|nodes$id==11370|nodes$id==10305|nodes$id==10319|nodes$id==9918|
              nodes$id==9106|nodes$id==8091|nodes$id==7452]<-2
nodes$level[is.na(nodes$level)]<-4

```  

In order to customize the plot I have included `level` and `font.size` in the nodes dataframe; the level assignment orders the nodes such that the U.S. node is at the top of the plot and the Mexico node is at the bottom of the plot. Years will be listed in the middle. I've used the `visHierarchicalLayout` command to accomplish this customization.  
Additionally, colors are selected via clustering and "betweenness", which is defined as the importance of each node to the graph. The greater the number of connections to a node the higher the index of "betweenness"; this is calculated using `igraph` and merged with the dataframe,then used to determine color, group and node size in the plot.  

```{r message=FALSE, warning=FALSE}

#vertices are nodes, base data are edges.
gd<- graph.data.frame(edges, directed=FALSE, vertices=nodes)

#calculate betweenness and store as "value"
nodes$value <- betweenness(gd)

#determine cluster via betweenness
cluster <- cluster_edge_betweenness(gd)

#set group identity
nodes$group <- cluster$membership

#plot the result
visNetwork(nodes, edges, main = list(text = "Tuberculosis, U.S. and Mexico",
                                     style = "font-family:Calibri;font-weight:bold;font-size:22px;text-align:center;"),
           width = "100%")%>%
  visHierarchicalLayout(
    direction="UD",
    treeSpacing = 500
  )
```

#Query: South America  

To plot a few countries in South America I'll follow the same process.  

```{r message=FALSE,warning=FALSE}

#query for countries in South America; all cases are totaled in the last command.
query <- "MATCH (a:country)-->(b:year)
WHERE a.country = 'Argentina' 
OR a.country = 'Brazil' 
OR a.country = 'Chile'
OR a.country = 'Venezuela (Bolivarian Republic of)'
OR a.country = 'Peru'
OR a.country = 'Colombia'
AND a.country = b.country AND a.year = b.year
RETURN a.country AS country, SUM(DISTINCT a.adult+a.child+a.elderly) AS cases, b.year AS year"

#data frame stores results
SAcountry <- cypher(graphtb, query)

library(DT)
#view results
datatable(SAcountry)

```

#Plot: South America  

Venezuela's name is abbreviated to fit in the plot, and cases for each country are assigned a level via comparison with a list of values. Font sizes are adjusted.  

```{r message=FALSE,warning=FALSE}

edges1<-SAcountry[c("country","cases")]
colnames(edges1)<-c("from","to")
edges2<-SAcountry[c("cases","year")]
colnames(edges2)<-c("from","to")
edges<-data.frame(rbind(edges1,edges2))
edges$from[edges$from=="Venezuela (Bolivarian Republic of)"]<-"Venezuela"

#keep case totals from the countries we'd like to put at the top of the plot
caseslist<-SAcountry$cases[SAcountry$country=="Brazil" | SAcountry$country=="Chile" | SAcountry$country=="Peru"]
nodes<-data.frame(id=unique(c(SAcountry$country,SAcountry$year,SAcountry$cases)),
                  font.size=32)
nodes$id[nodes$id=="Venezuela (Bolivarian Republic of)"]<-"Venezuela"
nodes$label<-nodes$id
nodes$font.size<-ifelse(nodes$id=="Brazil" | nodes$id=="Chile" |
                          nodes$id=="Peru" | nodes$id=="Colombia" |
                          nodes$id=="Venezuela" | nodes$id=="Argentina",48,14)
nodes$level[nodes$id=="Brazil" | nodes$id=="Chile" | nodes$id=="Peru"]<-1

nodes$level[nodes$id>=1995 & nodes$id <= 2013]<-3
nodes$level[nodes$id=="Colombia" | nodes$id=="Venezuela" | nodes$id=="Argentina"]<-5

#assign levels to case totals of countries at the top of the plot
condition<-nodes$id %in% caseslist
nodes$level[condition]<-2

nodes$level[is.na(nodes$level)]<-4

#vertices are nodes, base data are edges.
gd<- graph.data.frame(edges, directed=FALSE, vertices=nodes)

#calculate betweenness and store as "value"
nodes$value <- betweenness(gd)

#determine cluster via betweenness
cluster <- cluster_edge_betweenness(gd)

#set group identity
nodes$group <- cluster$membership

#plot the result
visNetwork(nodes, edges, main = list(text = "Tuberculosis, South America",
                                     style = "font-family:Calibri;font-weight:bold;font-size:22px;text-align:center;"),
           width = "100%")%>%
  visHierarchicalLayout(
    direction="UD",
    treeSpacing = 500
  )

```  

#Advantages and Disadvantages  

In my opinion, the main advantages of NoSQL databases such as Neo4j are performance and scalability. According to popular opinion, NoSQL databases are better suited for big data, whether structured or unstructured; they make use of object-oriented programming and process queries more quickly. Additionally, NoSQL platforms tend to be open source. NoSQL will no doubt gain traction as it is adopted by more business organizations.